Kindly Bent to Free Us

### 2019-07-05 kindly/lang.tex
Reordering.

Environments are composed of value and type bindings, along with type
declarations. Type declarations, of the form
$\tydecl{t}{\kschm}{K}{\tau}$, introduce both a new type constructor $\T{t}$ and
a data constructor $K$.

\begin{figure*}[t]
  \centering
  \input{grammar}
  \caption{Syntax}
  \label{grammar}
\end{figure*}
\input{internal-grammar}

\input{fig-reduction}
\input{fig-code}

\clearpage
\subsection{Typing}
\label{sdtyping}

@@

\subsubsection{Regions and Borrows}

\begin{figure}[!h]
  \centering
  \begin{subfigure}{0.4\linewidth}
    \begin{mathpar}
      \ruleSDRegion
    \end{mathpar}
    \caption{The {\sc Region} rule}
  \end{subfigure}\hfill
  \begin{subfigure}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\Lleftarrow$ }
      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
      >{$}l<{$}
      r}

      (\kaff_n\lk k\lk\kaff_\infty)
      &\svar[\MBORROW]{x}{\tau}^n
      &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
      &Mut
    \end{tabular}
    \caption{Borrowing rules for bindings}
  \end{subfigure}
\end{figure}

\subsubsection{Bindings and Closures}

One essential aspect of linear types is to control which variables
are allowed to be unused. This is usually done through \emph{weakening},
which allow to discard bindings from the environment.
Another important feature is to control \emph{captures} done by closures.
In \lang, we enforce both properties through the use of environment-wide
kind inequalities, which we note $\Cleq{\E}{k}$.
The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
$\bvar{x}{\tau}$ in $\E$,
if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
then $\Cleq{k'}{k}$.
While this definition is incomplete, it is sufficient to understand
its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
%
The {\sc Var} rule looks up the type scheme of the variable $x$ in
the environment $\E$
and instantiate a concrete type through the substitution $\unif$. It then
checks that the other bindings in $\E$ can safely be discarded by
ensuring they are not linear. This is achieved through the constraint
$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
%
The {\sc Lam} constraints must ensures that if the body of the lambda captures
linear or affine variable, the content will be reflected in the kind
annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
If one of the binding of $\E$ constraints a type whose kind is affine,
we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
the kind on the arrow is at least affine.

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleSDLam

    \ruleSDVar
  \end{mathpar}
  \caption{The {\sc Lam} and {\sc Var} rules}
  \label{selectrules:binders}
\end{figure}

@@

\begin{figure}[!h]
  \centering
  \begin{subfigure}{0.35\linewidth}
    \begin{mathpar}
      \ruleSDPair
    \end{mathpar}
    \caption{The {\sc Pair} rule}
  \end{subfigure}\hfill
  \begin{subfigure}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\Lleftarrow$ }
      >{$}c<{$}@{ $=$ }
      >{$}c<{$}@{ $\ltimes$ }
      >{$}c<{$}r}

@@

@@

\clearpage

\subsection{Annotating regions}
\label{regionannot}

So far, all \lang programs have been fully annotated with regions information.
We now show how to infer these regions annotations based on
optionally-annotated programs.
First, we extend the region annotation to $\region{S}{E}$ where $S$ is
a set of variables. This annotation, defined below, is equivalent to nested
region annotations for each individual variable.

\begin{align*}
  \region{x;S}{e} &= \region{x}{\region{S}{e}}& \region{\emptyset}{e} &= e\\
\end{align*}

\Cref{fig:region-annotation} define a rewriting relation $\RannotT{e}{e'}$
which indicates that an optionally annotated term $e$ can be written
in a fully annotated term $e'$.
Through the rule \textsc{Rewrite-Top}, this is defined
in term of an inductively defined relation
$\Rannot{e}{e'}{S}$ where $n$ is the current nesting and $S$ is a set of
variable that are not enclosed in a region yet.
The base cases are constants, variables and borrows.
The general idea is to start from the leafs of the syntax tree, create a
region for each borrow, and enlarge the region as much as possible.
This is implemented by a depth-first walk of the syntax
tree which collects each variable that has a corresponding borrow.
At each step, it rewrites the inner subterms,
consider which borrow must be enclosed by a region now, and
return the others for later enclosing. Binders force immediate
enclosing of the bound variables, as demonstrated in rule \textsc{Rewrite-Lam}.
For nodes with multiple children, we
use a scope merge operator to decide if regions should be placed and where.
This is shown in rule \textsc{Rewrite-Pair}.
The merge operator, written $\getBorrows{B_l}{B_r}{(S_l,S,S_r)}$, takes
the sets $B_l$ and $B_r$ returned by rewriting the subterms
and returns three sets: $S_l$ and $S_r$ indicates the variables
that should be immediately enclosed by a region on the left and right
subterms and $S$ indicates the set of the yet-to-be-enclosed variables.
As an example, the rule \textsc{AnnotRegion-MutLeft} is applied
when there is an immutable borrow and a mutable borrow. In that case, a
region is created to enclose the immutable borrow, while the mutable
borrow is left to be closed later. This is coherent with the rules
for environment splitting and suspended bindings from \cref{sdtyping}.
%
Explicitly annotated regions are handled specially through
rule \textsc{Rewrite-Region}. In that case, we assume that all inner
borrows should be enclosed immediately.

\begin{figure*}[!hbt]
  \centering
  \input{region-annot}
  \caption{Automatic region annotation --- $\RannotT{e}{e'}$}
  \label{fig:region-annotation}
\end{figure*}

>>>>

% Environments are composed of value and type bindings, along with type
% declarations. Type declarations, of the form
% $\tydecl{t}{\kschm}{K}{\tau}$, introduce both a new type constructor $\T{t}$ and
% a data constructor $K$.

\begin{figure*}[t]
  \centering
  \input{grammar}
  \caption{Syntax}
  \label{grammar}
\end{figure*}

\subsection{Typing}
\label{sdtyping}

@@

\subsubsection{Environments and bindings}

One main feature of our system is to control the usage of variables, either
through
linearity and affinity, which controls how variables are shared or ignored,
or through borrows, which allow to circumvent the linearity rules
in a controlled way.
In order to check these properties, we introduce several new bindings
in our typing environments.
The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
On top of the traditional type and value bindings, noted respectively
$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
bindings.
Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ which indicates
that the borrow $\borrow{x}$ can be used directly, as implemented
by the rule {\sc Borrow} in \cref{selectrules:borrow}.
Suspended bindings, noted $\svar{x}{\schm}^n$, indicates that $x$
cannot be used immediately, but will be usable in a region
marked with $x$ (as detailed in \cref{sdtyp:regions}).

In \lang, many properties are encoded through the use of environment-wide
constraints. Here we present how to control both unused variables and
closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
$\bvar{x}{\tau}$ in $\E$,
if type $\tau$ has kind $k'$ ($\inferSK{C}{\E}{\tau}{k'}$)
then $\Cleq{k'}{k}$.
While this definition is incomplete, it is sufficient to understand
its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
%
The {\sc Var} rule looks up the type scheme of the variable $x$ in
the environment $\E$
and instantiate a concrete type through the substitution $\unif$. It then
checks that the other bindings in $\E$ can safely be discarded by
ensuring they are not linear. This is achieved through the constraint
$\Cleq{\E\Sdel{X}}{\kaff_\infty}$.
%
The {\sc Lam} constraints must ensures that if the body of the lambda captures
linear or affine variable, the content will be reflected in the kind
annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
If one of the binding of $\E$ constraints a type whose kind is affine,
we will derive the constraints $\Cleq{\kaff_n}{k}$, which ensures that
the kind on the arrow is at least affine.

\begin{figure}[!h]
  \begin{minipage}{0.33\linewidth}
    \begin{align*}
      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
      B &::=\ \bnone \tag{Empty} \\
         &\mid \bvar{x}{\schm}\tag{Variables}\\
         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
         &\mid \bvar{\tvar}{\kschm}\tag{Types}
         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
    \end{align*}
    \caption{Type environments}
    \label{grammar:env}
  \end{minipage}\hfill
  \begin{minipage}{0.65\linewidth}
    \begin{mathpar}
      \ruleSDBorrow
      \and
      \ruleSDLam
      \and
      \ruleSDVar
    \end{mathpar}
    \caption{The {\sc Borrow} rule}
    \label{selectrules:borrow}
  \end{minipage}
\end{figure}

@@

Linear and affine types also control the copy of variables.
We now showcase how this is achieved in \lang through the {\sc Pair} typing rule
in \cref{sdtyp:pair}.
\TODO{Explain splitting}

\begin{figure}[!h]
  \centering
  \begin{subfigure}{0.35\linewidth}
    \begin{mathpar}
      \ruleSDPair
    \end{mathpar}
    \caption{The {\sc Pair} rule}
    \label{sdtyp:pair}
  \end{subfigure}\hfill
  \begin{subfigure}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\vdash_e$ }
      >{$}c<{$}@{ $=$ }
      >{$}c<{$}@{ $\ltimes$ }
      >{$}c<{$}r}

@@

\subsubsection{Regions}
\label{sdtyp:regions}

In order to provide good support for imperative programming, \affe allows
to borrow resources.
Such borrows are constrained to a given syntactic region. We note
$\region{x}{e}$ for the region allowing borrows of $x$ to be used in $e$
and where $n$ is the nesting of the region.
Borrows can not be used outside of a corresponding region.
These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.

\TODO{Transition}

In order to use a suspended binding for the variable $x$,
we need to enter a \emph{region} associated with $x$,
noted $\region{x}{e}$.
The main idea of region typing (implemented by the {\sc Region} rule)
is to replace suspended bindings by borrow bindings, typecheck the body
of the region, and finally ensure that the borrow made available inside
the region does not leak outside.
This last check is done by using integer indices that correspond to the nesting
level of the region. The kind of borrow is annotated with the integer
corresponding to their region, and a final constraint ensure that
the return type of the region does not contain such borrow.

We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{env:rule:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ for our example and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

\begin{figure}[!h]
  \begin{minipage}{0.35\linewidth}
    \centering
    \begin{mathpar}
      \ruleSDRegion
    \end{mathpar}
    \caption{The {\sc Region} rule}
    \label{selectrules:region}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\vdash_e$ }
      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
      >{$}l<{$}
      r}

      (\kaff_n\lk k\lk\kaff_\infty)
      &\svar[\MBORROW]{x}{\tau}^n
      &\bvar{\borrow[\MBORROW]{x}}{\borrowty[\MBORROW] k{\tau}}
      &Mut
    \end{tabular}
    \caption{Borrowing rules for bindings -- $\lregion{C}{x}{\E}{\E'}$}
    \label{env:rule:borrow}
  \end{minipage}
\end{figure}

@@

\input{internal-grammar}
\input{fig-reduction}
\input{fig-code}
